Simulating High-Dimensional Multivariate Data

### 2021-02-25 030-bigsimr-package.Rmd
removed ref to cor_fast

As mentioned, the user must decide how to describe correlation, based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel. (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

>>>>

As mentioned, the user must decide how to describe correlation, based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

### 2021-02-25 050-examples.Rmd
removed ref to cor_fast

```{r ch050-readBRCA, cache=FALSE}
d <- 1000
brca1000 <- example_brca %>%
  select(all_of(1:d)) %>%
  mutate(across(everything(), as.double)) # cor_fast currently doesn't work with integers
```

@@

```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
# Estimate Spearman's correlation on the count data
type <- 'Spearman'
R_S <- bs$cor_fast(as.matrix(brca1000), bs[[type]])
```

@@

```{r ch050-makeFigBRCA, cache=FALSE}
if(!file.exists("fig/ch050-figBRCA.png")) {
    ## estimate rho from sims and plot
    R_S_hat <- cor_fast(sim_nbinom, method = type)
    ## triplot
    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
    par(mfrow=c(1,3))
    plot(R_S[lower.tri(R_S)], R_S_hat[lower.tri(R_S_hat)],
         xlim = c(-1, 1), ylim = c(-1, 1),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
         xlab = "Target Correlation", ylab = "Estimated Correlation")
         ## main = "Spearman Cor. Matching")
    abline(a = 0, b = 1, col="red", lty="dashed")
    ## estimate prob,size from sims and plot
    nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
    plot(sizes, nbinom_size_prob_hat[1,],
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Size", ylab = "Estimated Size")
         ## main = "Marginal NB Size Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    plot(log(probs), log(nbinom_size_prob_hat[2,]),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
         ## main = "Marginal NB Prob Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    dev.off()
    par(mfrow=c(1,1))
}
```

@@

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
## Simulate random vectors equal to the sample size
N <- nrow(brca)
## create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs::cor_fast(tmpSim, bs[[type]])
}, simplify = FALSE)
## find quadratic loss at each rep
quadLoss <- sapply(simRho, rags2ridges::loss, T = R_S, type = "quadratic")
```

```{r ch050-lossResults, cache=FALSE}
frobeniusRDS <- paste0("results/brca", d, "_frobenius.rds")
if (!file.exists(frobeniusRDS)) {
    N <- nrow(brca1000)
    ## create m random vectors and estimate correlation
    simRho <- replicate(n = m, expr = {
        tmpSim <- bs$rvec(N , R_X, nb_margins)
        bs$cor_fast(tmpSim, bs[[type]])
    }, simplify = FALSE)
    frobenius <- sapply(simRho, rags2ridges::loss, T = R_S, type = "frobenius")
    saveRDS(frobenius, frobeniusRDS)
}
frobenius <- readRDS(frobeniusRDS)
```

>>>>

```{r ch050-readBRCA, cache=FALSE}
d <- 1000
brca1000 <- example_brca %>%
  select(all_of(1:d)) %>%
  mutate(across(everything(), as.double))
```

@@

```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
# Estimate Spearman's correlation on the count data
type <- 'Spearman'
R_S <- bs$cor(as.matrix(brca1000), bs[[type]])
```

@@

```{r ch050-makeFigBRCA, cache=FALSE}
if(!file.exists("fig/ch050-figBRCA.png")) {
    ## estimate rho from sims and plot
    R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs[[type]])
    ## triplot
    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
    par(mfrow=c(1,3))
    plot(R_S[lower.tri(R_S)], R_S_hat[lower.tri(R_S_hat)],
         xlim = c(-1, 1), ylim = c(-1, 1),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
         xlab = "Target Correlation", ylab = "Estimated Correlation")
         ## main = "Spearman Cor. Matching")
    abline(a = 0, b = 1, col="red", lty="dashed")
    ## estimate prob,size from sims and plot
    nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
    plot(sizes, nbinom_size_prob_hat[1,],
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Size", ylab = "Estimated Size")
         ## main = "Marginal NB Size Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    plot(log(probs), log(nbinom_size_prob_hat[2,]),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
         ## main = "Marginal NB Prob Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    dev.off()
    par(mfrow=c(1,1))
}
```

@@

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
## Simulate random vectors equal to the sample size
N <- nrow(brca)
## create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs[[type]])
}, simplify = FALSE)
## find quadratic loss at each rep
quadLoss <- sapply(simRho, rags2ridges::loss, T = R_S, type = "quadratic")
```

```{r ch050-lossResults, cache=FALSE}
frobeniusRDS <- paste0("results/brca", d, "_frobenius.rds")
if (!file.exists(frobeniusRDS)) {
    N <- nrow(brca1000)
    ## create m random vectors and estimate correlation
    simRho <- replicate(n = m, expr = {
        tmpSim <- bs$rvec(N , R_X, nb_margins)
        bs$cor(tmpSim, bs[[type]])
    }, simplify = FALSE)
    frobenius <- sapply(simRho, rags2ridges::loss, T = R_S, type = "frobenius")
    saveRDS(frobenius, frobeniusRDS)
}
frobenius <- readRDS(frobeniusRDS)
```
